---
title: "Wasserstein Distance calculation"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{wasserstein_metric}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```


## Mathematical definition

The Wasserstein distance (sometimes also: Wasserstein metric, earth mover
distance [EMD], transport distance) is a metric to describe the distance
between two continuous distributions.

To compare two distributions $A$ and $B$ we make use of the $L^2$ Wasserstein
distance $W$ in the representation
$$W := W(F_A, F_B) 
= \bigg( \int_0^1 \big|F^{-1}(u) - F^{-1}(u) \big|^2 du \bigg)^\frac{1}{2}$$
with $F_A$ and $F_B$ as the cumulative distribution functions (CDFs) of A and
B.

The squared $L^2$ Wasserstein distance $d := W^2$ is of special interest to us,
as it offers the following decomposition into location, shape, and size terms:
$$d := d(F_A, F_B) 
= \int_0^1 \big|F^{-1}(u) - F^{-1}(u) \big|^2 du 
= \underbrace{\big(\mu_A - \mu_B\big)^2}_{\text{location}} 
+ \underbrace{\big(\sigma_A - \sigma_B\big)^2}_{\text{size}} 
+ \underbrace{2\sigma_A \sigma_B \big(1 - \rho^{A,B}\big)}_{\text{shape}}$$

with $\mu_A, \mu_B$ the means, $\sigma_A, \sigma_B$ the standard deviations,
and $\rho^{A,B}$ the Pearson correlation of $A$ and $B$.


## Approximation on Discrete Distributions

For discrete empirical distributions $\hat{A}$ and $\hat{B}$ we will instead
model the distributions with the empirical CDFs $\hat{F}_A$ and $\hat{F}_B$. 
The squared $L^2$ Wasserstein distance $d$ is the approximated as the mean
difference between $K$ empirical equidistant quantiles $\alpha_k$ of the
quantile functions $Q_A$ and $Q_B$ of $\hat{A}$ and $\hat{B}$.
$$d(\hat{F}_A, \hat{F}_B) 
\approx \frac{1}{K} \sum_{k=1}^K \big(Q_A^{\alpha_k} - Q_B^{\alpha_k} \big)
\approx \big(\hat{\mu}_A - \hat{\mu}_B\big)^2 
+  \big(\hat{\sigma}_A - \hat{\sigma}_B\big)^2
+ 2\hat{\sigma}_A \hat{\sigma}_B \big(1 - \hat{\rho}^{A,B}\big)$$

## Three Implementations


The package `waddR` offers functions for an exact computation and the two
approximations explained above.

We will use pseudo-random normal distributed samples to show all three
functions.
```{r setup}
library(waddR)

set.seed(24)
x <- rnorm(100, mean = 42, sd = 2)
y <- rnorm(100, mean = 38, sd = 2)
```

The most exact computation is a faster reimplementation in Cpp of the
`wasserstein1d` function from the package `transport`.
```{r wassersein_exact}
wasserstein_metric(x, y)
```

The other two implementations are approximation based on the formula above.
The first one is `squared_wass_approx` and approximates the squared $L^2$
Wasserstein distance by calculating the mean squared difference of the
empirical equidistant quantile functions of the two input vectors.
By default, 1000 quantiles are used.
```{r wassersein_approx}
squared_wass_approx(x, y)
```

The function `squared_wass_decomp` approximates the squared $L^2$ Wasserstein
distance by calculating and adding the decomposites location, size, and shape.
It also returns the decomposite values.
```{r wassersein_decomp}
squared_wass_decomp(x, y)
```



## See Also

* [Two-sample test](wasserstein_test.html) based on a decomposition of the
Wasserstein distance between two distributions to check for differences

* Detect and describe 
[differential expression in scRNAseq data](wasserstein_singlecell.html)
