---
title: "Detecting Differential Expression in Single-Cell RNAseq data"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{wasserstein_singlecell}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```



With [`waddR`](waddR.html) we present an alternative procedure to identify
differential distributions in single-cell RNAseq data, that is capable of 
characterizing different types of differential distribution.
Here, we will show that we are able to distinguish differences between two 
conditions with regard to location, shape and size.

## Application to an Example Data Set from Conquer Database

This an example on how to analyse the expression of 18 immune cell in two 
hypothetical conditions.
For an example dataset we use GSE41265, which can be downloaded from the
[conquer database](http://imlspenticton.uzh.ch:3838/conquer/).
To read it, install the packages `MultiAssayExperiment` and 
`SummarizedExperiment` from BioConductor.

(For a more detailed description of how to handle datasets from the conquer
database, see the 
[conquer R tutorial](https://github.com/markrobinsonuzh/conquer/blob/master/shiny-download/tutorial.md).)

First, load all required libraries and the example .rds data file.

```{r}
library(waddR)
suppressPackageStartupMessages(library(SummarizedExperiment))
suppressPackageStartupMessages(library(MultiAssayExperiment))

example.dataset <- "GSE41265.rds"
if (!file.exists(example.dataset)) {
    download.file(
    "http://imlspenticton.uzh.ch/robinson_lab/conquer/data-mae/GSE41265.rds",
    destfile = example.dataset)
}
gse41265 <- readRDS(example.dataset)
```

Extract gene abundance from the dataset
```{r}
gene_abd <- experiments(gse41265)[[1]]
```

We are interested in looking at the "Transcripts per Million" matrix, which we
will interprete as gene expression levels. 
The `wasserstein.sc` function provided in `waddR` can only consider one value
per sample in a distribution and gene -- the individual gene expression levels.
```{r}
data <- assays(gene_abd)[["TPM"]]
```

For this demonstration we randomly select 100 out of the 45,686 transcripts and
define arbitrary conditions that we will compare in the `wasserstein.sc` test.
```{r}
rand_rows <- sort(sample(dim(data)[1], 100))
data <- data[rand_rows, ]
conditions <- c(rep(1,10), rep(2,8))
```

First we do a one-step test of all transcripts in the two conditions with
p-values based on one thousand permutations
```{r}
onestep <- "OS"
permutations <- 1000
testresults.os <- wasserstein.sc(data, conditions, permutations, onestep)
```

Seeing, that a lot of zero expression levels are present int the data we repeat
our test, looking at zero and non-zero expression levels separately this time
```{r}
twostep <- "TS"
permutations <- 1000
testresults.ts <- wasserstein.sc(data, conditions, permutations, twostep)
```

For every gene, a test result is printed, describing how the distribution
differs in size,
shape and location between the two conditions, while providing p-values and
errors.
```{r}
head(testresults.ts)

```

## See Also

* Fast and accurate 
[calculation of the  Wasserstein distance](wasserstein_metric.html)

* [Two-sample test](wasserstein_test.html) to check for differences between two
distributions

