---
title: "The Two-Sample Wasserstein Test"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{waddR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Test for Differences Between Two Distributions 

The package `waddR` offers a generalized testing procedure to detect if two samples $A$ and $B$ (e.g. from two different conditions) are differentially distributed.
We are essentially testing the null hypothesis $\mathcal{H}_0: F_A = F_B$ against $\mathcal{H}_1: F_A \neq F_B$.


## Showcasing the Comparison of Distributions

For showcasing the capabilities of this testing procedure we will compare a vector of 500 elements drawn from a normal $\mathcal{N}(30,2)$ distribution with four different vectors of around 500 elements.

The first vector we are comparing to is also $\mathcal{N}(30,2)$ distributed, but only 498 elements in size.
The other three vectors differ in mean, standard deviation, and shape and should show up as differentially distributed.

```{r setup}
library(waddR)

x <- rnorm(500, 30, 2)
same.dist1 <- rnorm(498, 30,2)
diff.dist1 <- rnorm(503, 30, 10)
diff.dist2 <- rnorm(511, 20, 2)
diff.dist3 <- c(rnorm(234, 27, 2), rnorm(260, 33, 2))
```

Comparing the two observations drawn from the same distribution yields a rather high p-value thus indicating that $\mathcal{H}_0$ can not be rejected.

```{r test_eq}
wasserstein.test(x, same.dist1, method = "SP")[c("pval","location","shape","size")]
```

For the two distributions that differ in location (`diff.dist2`) and size (`diff.dist1`), we obtain very low p-values indicating significant differential distribution.
We can confidently reject $\mathcal{H}_0$ in favor of $\mathcal{H}_1$

```{r test_diff1}
# diff.dist1 differs from x in its size (standard deviation)
wasserstein.test(x, diff.dist1, method = "SP")[c("pval","location","shape","size")]

# diff.dist2 differs from x in its location (mean)
wasserstein.test(x, diff.dist2, method = "Sp")[c("pval","location","shape","size")]
```

For the third distribution that we designed to be differentially distributed, we also obtain a low enough p-value to reject  $\mathcal{H}_0$ in favor of $\mathcal{H}_1$ 

```{r test_diff3}
# diff.dist3 differs from x in its shape
wasserstein.test(x, diff.dist3, method = "SP")[c("pval","location","shape","size")]
```



## See Also

* Fast and accurate [calculation of the  Wasserstein distance](wasserstein_metric.html)

* Detect and describe [differential expression in single-cell RNA sequencing data](wasserstein_singlecell.html)
